Unknown

### 2025-06-24 ms/ms.tex
updated bib and aa_stats

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\KOMAoption{captions}{tableheading}
\usepackage{hyperref}
\addtokomafont{disposition}{\rmfamily}
\usepackage{siunitx}
\newcommand{\aax}{$\mathrm{AA}$}
\newcommand{\Aamphi}{$A_{\mathrm{amphi}}$}
\newcommand{\Ahypo}{$A_{\mathrm{hypo}}$}
\newcommand{\agcurve}{$A \textendash g_\text{sw}$}
\newcommand{\aqcurve}{$A \textendash Q$}
\newcommand{\ca}{$C_\mathrm{a}$}
\newcommand{\caequals}[1]{$C_\mathrm{a} = \qty{#1}{\micro\mol\raiseto{-1}\mol}$}
\newcommand{\cabetween}[2]{#1 < $C_\mathrm{a} < \qty{#2}{\micro\mol\raiseto{-1}\mol}$}
\newcommand{\gsw}{$g_\text{sw}$}
\newcommand{\loggsw}{$\log(g_\text{sw})$}
\newcommand{\logA}{$\log(A)$}
\newcommand{\ppfd}{$\mathrm{PPFD}$}
\newcommand{\ppfdequals}[1]{$\mathrm{PPFD} = \qty{#1}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}$}
\newcommand{\ppfdqty}[1]{$\qty{#1}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}$}
\newcommand{\rh}{$\mathrm{RH}$}
\newcommand{\rhequals}[1]{$\mathrm{RH} = #1\%$}
\newcommand{\tleafequals}[1]{$T_\mathrm{leaf} = \qty{#1}{\degreeCelsius}$}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

@@

definite: Wei Shen Lim, Dachuan Wang, Chris Muir question: Sam McKlin,
Tom Buckley

\section{Main}\label{main}

{[}introduction{]} Stomata are important

Stomatal ratio is important unsolved problem

Amphi leaves are more common in sunny habitats and confer a great
benefit. (this pattern is also true in Solanum? could we analyze
herbarium data here?)

{[}maybe make distinction between amphi leaves and var in SR here since
all tomato species are amphi{]}

Leaves with greater stomatal density ratio are more in open, sunny
habitats because they deliver the greatest benefit in those
circumstances. An amphistomatous leaf increases photosynthetic carbon
gain compared to an otherwise identical hypostomatous leaf by increasing
conductance through the leaf intercellular airspaces and boudary layers;
the additional water loss through a second boundary layer is typically
small {[}cites{]}. We quantify this benefit as the amphistomy advantage
(AA = log(Aamphi/Ahypo). Why would AA be greater in sun than shade?
There are three nonmutually exclusive hypotheses that we classify as
`acclimatory', `plastic', and `constitutive'.

Acclimatory hypothesis: Leaves acclimated to high light intensity
typically increase total leaf stomatal conductance (increased CO\$\_2\$
supply) and upregulate Rubisco activation (increased CO\$\emph{2\$}
demand). A one-dimensional circuit model using the Farquhar-von
Caemmerer-Berry biochemical model of C\$\_3\$ photosynthesis shows that
both increased stomatal conductance and Rubisco activity should increase
AA, all else being equal (Supporting Information). If the acclimatory
hypothesis is correct, we predict that AAhigh \textgreater{} AAlow for
all species regardless of native habitat or growth environment. Plants
adapted to sunny, open habitats will evolve greater stomatal density
ratio to take advantage of regular exposure to high light intensity.

Plastic hypothesis: Individuals of the same genotype often develop
dramatically different leaves in sun and shade conditions (cite).
Plastic responses are likely adaptations to optimize photosynthesis at
different light intensities (cites), but changes in leaf anatomy and
biochemistry could modulate AA as a byproduct. Thicker or less porous
leaves, both of which are associated with high leaf mass per area (LMA),
will have lower g\_ias; leaves with increased total stomatal density and
Rubisco concentration have greater potential CO2 supply and demand.
Under the plastic hypothesis, we predict that AAsun \textgreater{}
AAshade for all species and light intensities. AAsun and gsmax,sun
should be positively associated with native light habitat.
{[}transition{]} We assume that genotypes adapted to sunny, open
habitats will express a phenotype best adapted to that environment when
leaves develop under high light intensity; genotypes adapted to shaded
closed habitats may be plastic, but suboptimal for light intensities
they do not regularly experience in nature.

{[}conceptual figure could also show differential benefit of amphi
leaves in sun/shade{]}

to plants growing there. in those It has been hypothesized that
amphistomy increases photosynthesis more in sunny places. Three
nonmutually exclusive adaptive explanations for why amphi leaves:

Conceptual figure explaining each hypothesis

\begin{itemize}
\tightlist
\item
  acclimation: greater demand, higher gs increase AA
\item
  developmental plasticity: light-induced changes in leaf anatomy
  modulate AA
\item
  constitutive: genetic differences in leaves adapted to different light
  habitats
\end{itemize}

We distinguished among these hypotheses by comparing AA among wild
tomato species from different native light habitats, grown under
simulated sun and shade light treatments, and measured under contrasting
light intensity (Figure of hypotheses and predictions). We measured AA
on 600 individual plants from 30 accessions (average of 10 replicates
per light treatment) using a recently developed method (\emph{1}). With
this method, we directly compare the photosynthetic rate of an untreated
amphistomatous leaf to that of the same leaf with gas exchange blocked
through the adaxial (upper) surface by transparent plastic, which we
refer to as `pseudohypostomy'. To compare amphi- and pseudohypostomatous
leaves at identical whole-leaf \(g_\text{sc}\), we measure \(A\) over a
range of \(g_\text{sc}\), inducing stomatal opening and closure by
modulating humidity (see Materials and Methods for further details).

Table of directional predictions (with table summarizing results? part
of conceptual figure? in supplement?)

caption: Directional predictions associated with each hypothesis
explaining why amphistomy advantage (AA) might be greater for leaves in
sunny, open habitats. For each hypothesis, we make predictions for how
native light habitat, light treatment, and light intensity would affect
AA.

\begin{longtable}[]{@{}
  >{\centering\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.1972}}
  >{\centering\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.3099}}
  >{\centering\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2535}}
  >{\centering\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2394}}@{}}
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\centering
hypothesis
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
native light habitat
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
light treatment
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
light intensity
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
acclimatory & cor(PAR, AA) = 0 & AAsun = AAshade & AA2000 \textgreater{}
AA150 \\
plastic & cor(PAR, AA) = 0 & AAsun \textgreater{} AAshade & AA2000 =
AA150 \\
constitutive & cor(PAR, AA) \textgreater{} 0 & AAsun = AAshade & AA2000
= AA150 \\
\end{longtable}

{[}add rows for when multiple hypotheses are supported simultaneously?
or put that in SI?{]}

{[}results{]} Amphistomy increases \(A\) in all accessions, in both sun
and shade leaves, and light intensities. We infer this from the fact
that blocking gas exchange in pseudohypostomatous leaves reduced \(A\)
by X-X\% depending on the accession, light treatment, and light
intensity (Table/figure). The AA is equivalent to an X-X\% change in
total \(g_\text{sc}\) (see SI section gs equivalency). But whereas
increasing \(g_\text{sc}\) would increase water loss as a necessary
by-product, amphistomy can increase \(A\) without any appreciable affect
on transpiration.

Sun leaves from high light habitats {[}not sure if htis result is true
yet{]} benefit the most from amphistomy because of a both developmental
plasticity and constitutive differences among accessions. {[}quantify
difference in AA and contribution of different affects{]}. Surprisingly,
light intensity had a little effect\ldots{} (should this be in this
paragraph, or it's own?).

{[}discussion{]} - sun/shade has long been appreciated and this shows
new trait that should be considered and that CO2 diffusion becomes major
limitation

\section{Materials and Methods}\label{materials-and-methods}

{[}this will be moved to SI eventually{]}

@@

We compared AA among 30 ecologically diverse accessions of wild tomato,
including representatives of all described species of \emph{Solanum}
sect. \emph{Lycopersicon} and sect. \emph{Lycopersicoides} (\emph{2})
and the cultivated tomato \emph{S. lycopersicum} var.
\emph{lycopersicum} \autoref{tab:accessions}. Due to constraints on
growth space and time, we spread out measurements over 80 weeks from May
1, 2022 to October 31, 2023. Replicates within accession were evenly
spread out over this period to prevent confounding of temporal variation
in growth conditions with accession. {[}anything else to say here? maybe
explain accession selection and phylogeny?{]}

\caption{\label{tbl-accessions}Solanum accessions}

>>>>

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\KOMAoption{captions}{tableheading}
\usepackage{hyperref}
\addtokomafont{disposition}{\rmfamily}
\usepackage{siunitx}
\newcommand{\aax}{$\mathrm{AA}$}
\newcommand{\Aamphi}{$A_{\mathrm{amphi}}$}
\newcommand{\Ahypo}{$A_{\mathrm{hypo}}$}
\newcommand{\agcurve}{$A \textendash g_\text{sw}$}
\newcommand{\aqcurve}{$A \textendash Q$}
\newcommand{\ca}{$C_\mathrm{a}$}
\newcommand{\caequals}[1]{$C_\mathrm{a} = \qty{#1}{\micro\mol\raiseto{-1}\mol}$}
\newcommand{\cabetween}[2]{#1 < $C_\mathrm{a} < \qty{#2}{\micro\mol\raiseto{-1}\mol}$}
\newcommand{\gmaxratio}{$g_\text{max,ratio}$}
\newcommand{\gsw}{$g_\text{sw}$}
\newcommand{\loggsw}{$\log(g_\text{sw})$}
\newcommand{\logA}{$\log(A)$}
\newcommand{\ppfd}{$\mathrm{PPFD}$}
\newcommand{\ppfdequals}[1]{$\mathrm{PPFD} = \qty{#1}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}$}
\newcommand{\ppfdqty}[1]{$\qty{#1}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}$}
\newcommand{\rh}{$\mathrm{RH}$}
\newcommand{\rhequals}[1]{$\mathrm{RH} = #1\%$}
\newcommand{\tleafequals}[1]{$T_\mathrm{leaf} = \qty{#1}{\degreeCelsius}$}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

@@

definite: Wei Shen Lim, Dachuan Wang, Chris Muir

question: Sam McKlin, Tom Buckley

I am using this approach to drafting the paper:

https://blogs.nature.com/nyc/2011/08/10/how-to-write-a-paper-one-possible-answer

\section{Punchline}\label{punchline}

Plastic changes in the structure of sun leaves can increase the
intercellular CO\(_2\) diffusion path length. Gas exchange through
stomata on both upper and lower leaf surfaces, amphistomy, may be a
compensatory mechanism to reduce the impact of increased diffusion path
length on photosynthesis without increasing water loss. We demonstrate
for the first time that sun leaves benefit most from amphistomy because
it compensates for plastic structural changes in leaf anatomy.

\section{Figures}\label{figures}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  competing hypotheses and predictions (or predictions after methods?)
  for the benefit of amphistomy
\item
  design of experiment to test competing hypotheses (with actual \#'s of
  replicate, curves, etc.)
\item
  Results on acclimatory, plastic, constitutive hypotheses
\item
  Additional supporting figure on change in LMA, reduction of Amass, and
  relationship between LMA and AA? Modeling?
\end{enumerate}

\section{Abstract}\label{abstract}

(use for GRC)

Developmental plasticity to light intensity modulates amphistomy
advantage

The presence of stomata on both leaf surfaces (amphistomy) can increase
photosynthesis in C\(_3\) plants by reducing the path length for
CO\(_2\) diffusion between substomatal cavities and chloroplasts.
Amphistomatous leaves are most common among herbaceous plants growing in
sunny habitats, including many crop species. The distribution of
amphistomatous leaves in nature may result from an increased
photosynthetic benefit of amphistomy under high light intensity, either
because of acclimatory, plastic, or constitutive variation in CO\(_2\)
supply or demand. We used a recently developed method to quantify
amphistomy advantage, the photosynthetic rate of an amphistomatous leaf
relative to an otherwise identical hypostomatous leaf, in 29 diverse
populations representing 15 species of wild tomatoes (\emph{Solanum}
sect. \emph{Lycopersicon} and sect. \emph{Lycopersicoides}). Plant grown
under high light intensity benefit more from amphostomy than those grown
under low light, regardless of acclimation or native light intensity.
Curvature of light response curves indicates that high-light leaves
benefit more from amphistomy because of greater resistance to CO\(_2\)
diffusion within the mesophyll. Developmental plasticity in leaf
thickness and/or cell arrangement to optimize light interception likely
increases under varying light intensity modulates CO\(_2\) diffusion
within the leaf and, hence, amphistomy advantage as a byproduct.
Developmental plasticity may therefore play an important and previously
unnoticed role in explaining the adaptive significance of amphistomy and
the distribution of amphistomatous leaves in nature.

What's surprising: role of developmental plasticity in explaining
comparative results; diffusional limitations through palisade ias is
strong and needs to be considered with light gradients in understanding
leaf strucuture- funciton

\section{Intro}\label{intro}

Stomata are microscopic pores on the surfaces of leaves and other
photosynthetic organs formed by a pair of guard cells. They are
essential for balancing carbon gained per unit water lost and permitted
vascular plants to grow tall on land by enabling access to CO\(_2\) for
photosynthetic carbon assimilation while preventing hydraulic failure in
variable environments (\emph{1}--\emph{3}). Optimal stomatal function
depends on both dynamic changes in aperture on the scale of minutes to
hours, as well as static anatomy determined by developmental plasticity
and constitutive genetic differences (\emph{4}--\emph{6}). Understanding
how stomata respond to environmental change over daily, developmental,
and evolutionary time is important for understanding adaptation
(\emph{1}, \emph{7}--\emph{12}), predicting paleoclimate from fossil
cuticles (\emph{13}--\emph{15}), and improving crops (\emph{16}).
Stomatal function contributes to global carbon and water cycles
(\emph{17}) and therefore predicting future climate (\emph{18}).

Despite extensive theoretical and empirical progress understanding
stomata function and anatomy from molecular to ecosystem levels, the
adaptive significance of amphistomatous leaves remains an important
unsolved problem in leaf structure-function relationships
(\emph{19}--\emph{26}). Amphistomatous leaves develop abaxial and
adaxial stomata whose aperture can be independently regulated
(\emph{27}--\emph{31}) to control gas exchange through each surface. All
else being equal, gas exchange through stomata on both surfaces
increases CO\(_2\) supply to chloroplasts by providing a second parallel
pathway through leaf intercellular airspaces, enhancing photosynthesis
(\emph{20}, \emph{32}). The extent to which amphistomy increases
CO\(_2\) supply depends on resistance to diffusion in intercellular
airspaces. This resistance can be low in thin, porous, amphistomatous
leaves (\emph{28}, \emph{33}), but may be more substantial in thick,
dense, hypostomatous leaves (\emph{34}). Amphistomatous leaves also lose
more water through evaporation because of a second boundary layer
conductance (\emph{35}), but the additional carbon gain should be enough
to offset this cost in most realistic scenarios (\emph{36}).

The paradoxical fact is that, despite the photosynthetic benefit, most
leaves are not amphistomatous. Many vertically oriented and/or
isobilateral leaves are amphistomatous (\emph{25}). But among
dorsiventral leaves, herbaceous plants in open, high light habitats tend
to have have amphistomatous leaves (\emph{22}, \emph{39}--\emph{44}).
Most other leaves, except those from aquatic habitats, are
hypostomatous, producing stomata only on the lower, abaxial surface.
Even resupinate leaves develop stomata on the lower, albeit adaxial
surface (\emph{45}), suggesting that leaf orientation (lower vs.~upper)
rather than leaf polarity (abaxial vs.~adaxial) is causal. The
covariation between stomatal density ratio and light habitat is both
qualitative and quantitative. A higher proportion of sun leaves are
amphistomatous (\textbf{salisbury\_causes\_1928?}) and the proportion of
stomata on the upper, adaxial surface increases with light (\emph{42},
\emph{43}). Resolving why high light intensity favors amphistomatous
dorsivental leaves is an important first step toward understanding
variation in stomatal density ratio and leaf structure-function
relationships more generally.

The overarching hypothesis is that leaves with greater stomatal density
ratio are more common in open, sunny habitats because they increase
photosynthesis most in those circumstances. An amphistomatous leaf
increases photosynthetic carbon gain compared to an otherwise identical
hypostomatous leaf by increasing conductance through the leaf
intercellular airspaces and boudary layers; the additional water loss
through a second boundary layer is typically small (\emph{35}). We
quantify this benefit as the amphistomy advantage (AA =
log(Aamphi/Ahypo) (\emph{20}, \emph{46}). Why would AA be greater in sun
than shade? We consider three nonmutually exclusive hypotheses that we
classify as `acclimatory', `plastic', and `constitutive'.

Acclimatory hypothesis: Photosynthetic induction to high light intensity
typically involves increases in total leaf stomatal conductance
(increased CO\(_2\) supply), the concentration of active Rubisco, and
electron transport capacity (increased CO\(_2\) demand). A
one-dimensional circuit model using the Farquhar-von Caemmerer-Berry
biochemical model of C\(_3\) photosynthesis (\emph{47}) shows that both
increased stomatal conductance and Rubisco activity should increase AA,
all else being equal (Supporting Information). If the acclimatory
hypothesis is correct, we predict that AAhigh \textgreater{} AAlow for
all species regardless of native habitat or growth environment. Plants
adapted to sunny, open habitats will evolve greater stomatal density
ratio to take advantage of regular exposure to high light intensity.

Plastic hypothesis: Individuals of the same genotype often develop
dramatically different leaves in sun and shade conditions (\emph{48}).
Plastic responses are likely adaptations to optimize photosynthesis at
different light intensities in variable environments (\emph{49}).
Plastic changes in leaf anatomy and biochemistry could modulate AA as a
byproduct. Thicker or less porous leaves, both of which are associated
with high leaf mass per area (LMA), will have lower \(g_\mathrm{ias}\);
leaves with increased total stomatal density and photosynthetic capacity
have greater potential CO\(_2\) supply and demand. Under the plastic
hypothesis, we predict that AAsun \textgreater{} AAshade for all species
and light intensities. Secondarily, AAsun and gsmax,sun should also be
positively associated with native light habitat if genotypes adapted to
sunny, open habitats if they can express a phenotype best adapted to
that environment when leaves develop under high light intensity.
Genotypes adapted to shaded, closed habitats may be plastic, but limits
on the width of their reaction norms prevent them from developing traits
optimal for conditions they do not regularly experience in nature.

Constitutive hypothesis: In environments that are relatively constant or
where environmental change cannot be anticipated by a reliable cue,
natural selection will favor constitutive expression of optimal
phenotypes. We therefore predict genotypes from more sunny, open
habitats will have consistently greater AA under also measurement and
growth light intensities. For herbaceous plants, light intensity is
largely a function of the tree canopy (\emph{50}). Herbs growing in the
open will regularly experience high light intensity; herbs growing under
a forest canopy will often experience low light intensity.

The primary directional predictions for each hypothesis are summarized
in Table X; detailed predictions for results that would indicate support
for multiple hypotheses are in Table SX.

CONCEPTUAL FIGURE HERE

caption: Directional predictions associated with each hypothesis
explaining why amphistomy advantage (AA) might be greater for leaves in
sunny, open habitats. For each hypothesis, we make predictions for how
native plant area index (PAI), growth light treatment, and measurement
light intensity would affect AA.

\begin{longtable}[]{@{}
  >{\centering\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.1728}}
  >{\centering\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.3580}}
  >{\centering\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.3210}}
  >{\centering\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.1481}}@{}}
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\centering
hypothesis
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
measurement light intensity
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
growth light intensity
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
native PAI
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
acclimatory & AA2000 \textgreater{} AA150 & AAsun = AAshade & cor(PAI,
AA) = 0 \\
plastic & AA2000 = AA150 & AAsun \textgreater{} AAshade & cor(PAI, AA) =
0 \\
constitutive & AA2000 = AA150 & AAsun = AAshade & cor(PAI, AA)
\textgreater{} 0 \\
\end{longtable}

We tested these hypotheses by comparing AA among amphistomatous wild
tomato species (\emph{51}) from different native light habitats, grown
under simulated sun and shade light treatments, and measured under
contrasting light intensity (Figure of hypotheses and predictions). We
measured AA on 572 individual plants from 29 accessions (average of 9.86
replicates per light treatment) using a recently developed method
(\emph{46}). With this method, we directly compare the photosynthetic
rate of an untreated amphistomatous leaf to that of the same leaf with
gas exchange blocked through the adaxial (upper) surface by transparent
plastic, which we refer to as `pseudohypostomy'. To compare amphi- and
pseudohypostomatous leaves at identical whole-leaf \(g_\text{sc}\), we
measure \(A\) over a range of \(g_\text{sc}\), inducing stomatal opening
and closure by modulating humidity (see Materials and Methods for
further details). We estimated `amphistomy advantage' (\aax)
\emph{sensu} (\emph{20}), but with modifications previously described in
(\emph{46}) and here (Materials and Methods). The native light intensity
was represented by plant area index (PAI
\(\unit{\meter\squared\per\meter\squared}\)), estimated using a global
gridded X-m2 resolution data set derived from the Global Ecosystem
Dynamics Investigation {[}GEDI; (\emph{52}){]} and georeferenced
accession collection information from the Tomato Genetics Resource
Center. The growth light intensities were 761 (sun treatment) and 115
(shade treatment) while all other environment conditions were nearly
identical (see \textbf{?@sec-methods} for supporting detail). The high
and low measurements intensities were \ppfdequals{2000} (97.8:2.24
red:blue) and \ppfdequals{150} (87.0:13.0 red:blue), respectively.

\subsubsection{MARKER}\label{marker}

Consistent with biophysical theory of CO\(_2\) diffusion within leaves,
AA \textgreater{} 0 for all accessions (Figure X). AA varied
substantially between measurement light intensities, growth light
intensities, and among accessions (point to model comparison evidence).
Measured under high light intensity, AA was consistently greater for sun
plants. The average AA among accessions in the shade treatment was 0.039
(range: 0.012--0.109; 21 of 29 accessions significant); however, the
same accessions grown at high light intensity showed a mean AA of 0.050
(range: 0.024--0.120; 25 of 29 accessions significant). Contrary to the
predictions of the assimilatory hypothesis, AA was greater in all
accessions under low measurement light intensity for both sun and shade
grown plants. The effect of low light on AA was more pronounced in the
sun-grown plants, where AA was significantly greater under low
measurement light intensity in 28 of 29 accessions compared to to 11
accessions for shade-grown plants. The overall average AA of shade and
sun grown plants measured under low light intensity was 0.066 (range:
0.019--0.143; 27 of 29 accessions significant) and 0.098 (range:
0.051--0.175; 29 of 29 accessions significant), respectively. There was
a slight tendency for accessions from more closed habitats (greater
PAI), but AA varied widely among accessions from open habitats (low PAI)
regardless of growth and measurement light intensities (Figure X). The
pattern of AA across wild tomatoes strongly supports the plasticity
hypothesis, argues against the acclimatory hypothesis, and provides only
weak support for the constitutive hypothesis.

Not sure where this goes: We infer this from the fact that blocking gas
exchange in pseudohypostomatous leaves reduced \(A\) by X-X\% depending
on the accession, light treatment, and light intensity (Table/figure).
The AA is equivalent to an X-X\% change in total \(g_\text{sc}\) (see SI
section gs equivalency). But whereas increasing \(g_\text{sc}\) would
increase water loss as a necessary by-product, amphistomy can increase
\(A\) without any appreciable affect on transpiration.

These results suggest that developmental plasticity, as opposed to
acclimation or adaptation to open habitats, may explain the explain the
long-standing observation taht amphistomatous leaves are more common in
sunny habitats. Our results add to existing explanations by showing high
light intensity \emph{per se} does not increase the benefit of
amphistomy, but rather than anatomical and biochemical change associated
with higher light intensity modulate AA. We cannot yet attribute
specific changes wrought by plasticity, but correlational evidence
suggests that changes in leaf anatomy may be important. Species
responded consistently by increasing stomatal density, especially on the
adaxial surface, increasing stomatal density ratio. Sun leaves also had
greater gs, Amax. However, these traits were not associated with change
in AA (table/figures).

from Mott and Oleary: ``However, Jones and Slatyer (5) reported a higher
mesophyll resistance for CO, entering through the upper stomata than for
the lower, and the data of Vaclavik (12) appear to support this
conclusion''

\section{Outtro - I think this goes at the
end??}\label{outtro---i-think-this-goes-at-the-end}

It is commonly assumed in comparative studies that most trait variation
between species is constitutive and genetically determined, but our
results suggest that plastic responses to light are necessary to explain
why \gmaxratio{} increases with light habitat. Reaction norms to light
intensity among wild tomatoes indicate that accessions from low light
habitats express low \gmaxratio{} when grown in shade, whereas
accessions from high light habitats have constitutively higher
\gmaxratio{} (Fig. rxn-norm). The larger response to light intensity is
driven by greater plasticty in leaf structure in accessions from
low-light habitats (Fig/table - I haven't tested this yet, so this is a
placeholder.) We infer that these structural changes in sun leaves are
costly because they increase the diffusion path length for CO\(_2\) and
reduce photosynthesis. As a consequence, sun leaves benefit more from
amphistomy. This is the first demonstration that sun leaves benefit more
from amphistomy because it compensates for developmental plasticity in
leaf structure.

Two outstanding areas for further research are 1) the specific
anatomical changes, such as leaf thickness or mesophyll porosity, that
increase the diffusion path length for CO\(_2\) in sun leaves, and 2)
the costs of amphistomy that could explain plasticity in \gmaxratio{} in
accessions from low-light habitats. Resolving the first question will
require detailed anatomical measurements on a large number of shade and
sun leaves with concomitant estimates of \aax. The fact that accessions
from low-light habitats are more plastic suggests that costs of
amphistomy are greatest in these environments. (not finished yet)

Problems to bring up in discussion: - all our species were amphi, so
they may all live in PAI below threshold to favor amphi - other
microclimate diffs not considered - don't know specific anatomical
change

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{7}
\tightlist
\item
  What it all means
\end{enumerate}

\begin{itemize}
\tightlist
\item
  Our study reveals an important role for plasticity in explaining why
  amphi leaves are more common in sunny habitats
\item
  Confirms that leaf structure of sun leaves incurs costs that
  amphistomy can mitigate
\item
  It will be useful to consider coordination between leaf anatomy,
  stomatal density, and ratio in future studies of leaf
  structure-function and in crops (or something)
\end{itemize}

\section{Materials and Methods
{[}\#sec-methods{]}}\label{materials-and-methods-sec-methods}

{[}this will be moved to SI eventually. note that much of it also the
same in solanum-aq project, so might change to cite that paper?{]}

(not sure where this goes) * We acclimated the focal leaf to high light
(\ppfdequals{2000}) and high relative humidity (\rhequals{70}) until
\(A\) and \gsw{} reach their maximum. After that, we decreased \rh{} to
\(\approx 10\%\) to induce rapid stomatal closure without biochemical
down

@@

We compared AA among 29 ecologically diverse accessions of wild tomato,
including representatives of all described species of \emph{Solanum}
sect. \emph{Lycopersicon} and sect. \emph{Lycopersicoides} (\emph{51})
and the cultivated tomato \emph{S. lycopersicum} var.
\emph{lycopersicum} \autoref{tab:accessions}. Due to constraints on
growth space and time, we spread out measurements over 61.1 weeks from
August 29, 2022 to October 31, 2023. Replicates within accession were
evenly spread out over this period to prevent confounding of temporal
variation in growth conditions with accession. {[}anything else to say
here? maybe explain accession selection and phylogeny?{]}

\begin{longtable}{>{\raggedright\arraybackslash}p{3.25cm}lrrrr}

\caption{\label{tbl-accessions}Accessions of \emph{Solanum} used in this
study. The species name, accession number, collection latitude,
longitude, elevation, and daily solar radiation estimated using the
SPLASH algorithm (see `Climate data'). TGRC: Tomato Genetics Resource
Center; PPFD: Photosynthetic Photon Flux Density.}
